---
title: "allourhardwork"
format: html
---

## making the dates in format YYYY-MM-DD
library(lubridate)

hot_days$Start_Date <- ym(hot_days$Start_Date)

cold_days$Start_date <- ym(cold_days$Start_Date)


## reading in data tables 
library(readr) 

cold_days <- read_csv("data/cold_days.csv")
View(cold_days)

hot_days <- read_csv("data/hot_days.csv")
View(hot_days)

elevation <- read_csv("data/elevation.csv")
View(elevation)

high_temp <- read_csv("data/highest_temp.csv")
View(high_temp)

humid <- read_csv("data/humidity.csv")
View(humid)

low_temp <- read_csv("data/lowest_temp.csv")
View(low_temp)

nrmavg <- read_csv("data/nrmavg.csv")
View(nrmavg)

nrmmax <- read_csv("data/nrmmax.csv")
View(nrmmax)

nrmmin <- read_csv("data/nrmmin.csv")
View(nrmmin)

UHI <- read_csv("data/UHI.csv")
View(UHI)


## Bar Graph comparing Denvers elevation to New Orleans elevation
```{r}
library(readr)
library(dplyr)
library(ggplot2)

elev_data <- read.csv("data/elevation.csv")

elev_data$High.Feet <- as.numeric(gsub(",", "", elev_data$High.Feet))
elev_data$Low.Feet <- as.numeric(gsub(",", "", elev_data$Low.Feet))

two_cities <- elev_data %>%
  filter(City %in% c("DENVER", "NEW ORLEANS"))

ggplot(two_cities, aes(x = City, y = High.Feet)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "City", y = "Elevation (High Feet)", title = "City Elevation Comparison") +
  theme_minimal()
```
With Denver's elevation being significantly higher than New Orleans will give us a foundations to compare the extreme changes that happen at different elevations. 


## Bar Plot comparing UHI temperature differences 
```{r}
library(readr)
library(dplyr)
library(ggplot2)

UHI <- read.csv("data/UHI.csv")

# Clean column names
names(UHI) <- gsub(" ", "_", names(UHI))

# Filter for Denver and New Orleans
city_uhi <- UHI %>%
  filter(City %in% c("Denver", "New Orleans"))

# Select only the relevant columns (City and UHI temperature effect)
city_uhi <- city_uhi %>%
  select(City, Citywide.urban.heat.island.effect..rolled.up.from.census.tract.data..weighted.by.area..in.degrees.F.)

# Rename the UHI column for clarity
colnames(city_uhi)[2] <- "UHI_Effect_F"

# Plot the UHI temperature differences
ggplot(city_uhi, aes(x = City, y = UHI_Effect_F, fill = City)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = round(UHI_Effect_F, 1)), vjust = -0.5) +  # Add value labels above bars
  labs(
    title = "Urban Heat Island Effect: Denver vs New Orleans",
    x = "City",
    y = "UHI Temperature Difference (°F)"
  ) +
  scale_fill_manual(values = c("skyblue", "pink")) +
  theme_minimal() +
  theme(legend.position = "none")
```
This plot shows the UHI difference (in °F) for both cities. New Orleans's higher UHI is expected because of the low-elevation and humidity and the fact that it is pretty warm there year round. This creates a greater challenge for New Orleans when it comes to managing heat. Denver's UHI isnt nearly as bad because of the dry air and higher elevation. 



